---
title: "Mediation-lavaan"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mediation-lavaan}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(MICr)
library(lavaan)
library(knitr)  # for kable.
```

In this vignette, we lay out a very simplified mediation model as a means of understanding what the MIC does.
```{r, XYZExample}
XYZmodel <- lavaan("Y ~ .2*X +.6*Z
                    Z ~ .8*X")
```

The model's MIC looks like this:
```{r}
kable(MIC(XYZmodel), caption="Simple Mediation MIC")
```

Note that Z->Y and X->Z remain identical, since there is only one path present to follow each link.  However, X->Y is two paths: one direct and one indirect.  The MIC path between them is the combination of the two.  Also, because `lavaan` orders the manifest variables as Y, X, Z, the MIC does as well.

For ease of reporting, this can also be displayed as a MIC table.

```{r}
MICTable(XYZmodel, caption="Simple Mediation MIC Table")
```

The table format has the easy ability to select using traditional data frame selection operators.  For example, if we only care about the causal influences on Z or on Y:
```{r}
theTable <- MICTable(XYZmodel)
onZ <- theTable[theTable$to == "Z",]
onY <- theTable[theTable$to == "Y",]

kable(onZ, caption="Simple Mediation MIC: Effects on Z", row.names=FALSE)
kable(onY, caption="Simple Mediation MIC: Effects on Y", row.names=FALSE)
```

This allows us to identify the differences in the causal structure in a more direct and sensible way.
