---
title: "SimpleMediation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SimpleMediation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
suppressPackageStartupMessages(library(MICr))
suppressPackageStartupMessages(library(OpenMx))
suppressPackageStartupMessages(library(knitr))  # for kable.
suppressPackageStartupMessages(library(DiagrammeR))
```

In this vignette, we lay out a very simplified mediation model as a means of understanding what the MIC does.
```{r, XYZExample}
XYZmodel <- mxModel("XYZ", manifestVars=c("X", "Y", "Z"), type="RAM",
  mxPath(from="Z", to="Y", values=.6, labels="\\Beta"),
  mxPath(from="X", to=c("Y", "Z"), values=c(.2, .8), labels=c("\\lambda", "\\alpha")),
  mxMatrix("Iden",3,name="I"))
```

The model's MIC looks like this:
```{r}
kable(MIC(XYZmodel), caption="Simple Mediation MIC")
```

Note that Z->Y and X->Z remain identical, since there is only one path present to follow each link.  However, X->Y is two paths: one direct and one indirect.  The MIC path between them is the combination of the two.

For ease of reporting, this can also be displayed as a MIC table.

```{r}
kable(MICTable(XYZmodel), caption="Simple Mediation MIC Table")
```

The table format has the easy ability to select using traditional data frame selection operators.  For example, if we only care about the causal influences on Z:
```{r}
theTable <- MICTable(XYZmodel)
onZ <- theTable[theTable$to == "Z",]
kable(onZ, caption="Simple Mediation MIC: Effects on Z", row.names=FALSE)
```

