---
title: "Matrices of Implied Causation in a Simple Mediation Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CLPMs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
In this vignette, we lay out a very simplified mediation model as a means of understanding how Matrices of Implied Causality (MICs) can be applied. We first need to load in the `MICr` package.  If `semPlot` is installed, we'll use it, too.

```{r setup, message=FALSE, warning=FALSE}
library(MICr)
library(OpenMx) 
if(library(semPlot, logical.return=TRUE))  drawPlots<-TRUE else drawPlots <- FALSE 
```

The primary model is a fairly straightforward mediation model.  X is the predictor, Z the mediator, and Y is the outcome.
```{r, XYZExample, message=FALSE, warning=FALSE}
XYZmodel <- mxModel("XYZ", manifestVars=c("X", "Y", "Z"), type="RAM",
  mxPath(from="Z", to="Y", values=.6, labels="\\Beta"),
  mxPath(from="X", to=c("Y", "Z"), values=c(.2, .8), labels=c("\\lambda", "\\alpha")),
  mxMatrix("Iden",3,name="I"))
if(drawPlots) semPaths(XYZmodel)
```

The model's MIC looks like this:
```{r}
MIC(XYZmodel)
```

Note that Z->Y and X->Z are identical to the path loadings, since there is only one path present to follow each link.  However, X->Y is two paths: one direct and one indirect.  The MIC path between them is the total effect--the combination of the influences across the two paths.

For ease of reporting, this can also be displayed as a MIC table.

```{r, results='asis'}
MICTable(XYZmodel, caption="Simple Mediation MIC Table")
```

The table format has the easy ability to select using traditional data frame selection operators, but the `MICTable` function also has helpers.  For example, if we only care about the causal influences on Z or on Y:
```{r, results='asis'}
MICTable(XYZmodel, caption="Simple Mediation MIC Table: Effects on Y", to="Y")
MICTable(XYZmodel, caption="Simple Mediation MIC Table: Effects on Z", to="Z")
```

This allows us to identify the unique elements of the causal structure in a more direct and sensible way.
